Using raw accelerometer data to predict high-impact mechanical loading

### 2022-02-27 manuscript/manuscript.Rmd
Revise the manuscript

But, even when knowing precisely which is the best loading regimen to improve bone health, the question of how to measure and monitor this load remains. Bone mechanical loading is usually measured in laboratory settings, using instruments such as force plates (FP) [@Medved2000]. Measurements in this instruments, that are confined to a laboratory space, are not applicable to controling the loads a person is routinely subjected in free-living contexts. To surpass this issue, in the last few years, methods using wearable devices such as accelerometers have been proposed to measure mechanical loading, more precisely,  ground reaction force (GRF) [@Neugebauer2012; @Neugebauer2014; @Fortune2014; @Neugebauer2018; @Veras2020] and loading rate (LR) [@Fortune2014; @Veras2020]. These devices are light, small, can be used throughout days, and have a relatively affordable price. The aforementioned characteristics make the accelerometers well suited instruments to measure mechanical loading on a daily basis.

However, the validity of the accelerometer-predicted mechanical loading values depends on some factors, as the sample characteristics [@Welk2005]. In fact, due to this, several calibration studies have been done in several populations, as children [@Neugebauer2012], adults [@Neugebauer2014; @Fortune2014], military personnel [@Neugebauer2018], and subjects with obesity [@Veras2020]. Another important factor that influence the predictions validity is the type of activities performed in the calibration study [@Bassett2012]. Until now, studies that calibrate accelerometers to predict mechanical loading variables have employed usually activities as walking [@Fortune2014; @Neugebauer2018; @Veras2020], or walking and running [@Neugebauer2012; @Neugebauer2014]. Nevertheless, activities eliciting higher impact loads have not yet been tested.

@@

The use of a prediction model based on accelerometer data should ideally be done in conditions similar to the study that validated the model, specially in terms of sample characteristics and type and intensity of the activities performed [@Welk2005]. Therefore, if we want to accurately predict the mechanical loading caused by high-impact activities, we should choose a prediction model that has been validated using these activities. Even though the cross-validation accuracy of such a model was poorer compared with other models developed for activities such as walking and running, as demonstrated by our results, the use of these models is crucial to achieve an acceptable prediction accuracy. It was demonstrated by testing a model previously developed by our group [@Veras2020] in a similar sample, but on walking, on the accelerometer-recorded jumping data from the current study. The accuracy indices yielded by applying these models have greatly worsened and these equations also produced some abnormal values, mainly in the pGRF model, in which it predicted negative values of force. These negative pGRF values were mainly as a result of a predictor, the second degree polynomial of the acceleration, possessing a negative coefficient. While this coefficient captured the plateauing of pGRF as pACC magnitude augments during walking, effectively improving the model [@Veras2020] for walking, this same pattern is not present in the pGRF vs. pACC relationship during jumping, and, therefore, only impairs the accuracy of these equations when applyed to this kind of activity.

@@

Although the models presented in this study showed to be valid, some limitations ought to be considered. First, the choice of activity to elicit high-impact loading was jumping, and several combinations of jump type and height were tested. Nevertheless, the results presented here cannot be assured for other high-impact activities, or even for other types of jump that were not tested. Second, no external sample for model validation was recruited. However, the leave-one-out cross-validation method that was used is the recommended in these situations [@Staudenmayer2012]. Third, to apply these models in accelerometer-recorded free-living physical activity data, there is a need to identify, in the acceleration signal, the periods in which the activity being performed is jumping. Therefore, methods for physical activity pattern detection should be used in combination with these models.

>>>>

But, even when knowing precisely which the best loading regimen is to improve bone health, the question of how to measure and monitor this load remains. Bone mechanical loading is usually measured in laboratory settings, using instruments such as force plates (FP) [@Medved2000]. Measurements in these instruments, that are confined to a laboratory space, are not applicable to controling the loads a person is routinely subjected in free-living contexts. To surpass this issue, in the last few years, methods using wearable devices such as accelerometers have been proposed to measure mechanical loading, more precisely,  ground reaction force (GRF) [@Neugebauer2012; @Neugebauer2014; @Fortune2014; @Neugebauer2018; @Veras2020] and loading rate (LR) [@Fortune2014; @Veras2020]. These devices are light, small, can be used throughout days, and have a relatively affordable price. The aforementioned characteristics make the accelerometers well suited instruments to measure mechanical loading on a daily basis.

However, the validity of the accelerometer-predicted mechanical loading values depends on some factors, as the sample characteristics [@Welk2005]. In fact, due to this, several calibration studies have been done in several populations, as children [@Neugebauer2012], adults [@Neugebauer2014; @Fortune2014], military personnel [@Neugebauer2018], and subjects with obesity [@Veras2020]. Another important factor that influences the predictions validity is the type of activities performed in the calibration study [@Bassett2012]. Until now, studies that calibrate accelerometers to predict mechanical loading variables have usually employed activities as walking [@Fortune2014; @Neugebauer2018; @Veras2020], or walking and running [@Neugebauer2012; @Neugebauer2014]. Nevertheless, activities eliciting higher impact loads have not yet been tested.

@@

The use of a prediction model based on accelerometer data should ideally be done in conditions similar to the study that validated the model, specially in terms of sample characteristics and type and intensity of the activities performed [@Welk2005]. Therefore, if we want to accurately predict the mechanical loading caused by high-impact activities, we should choose a prediction model that has been validated using these activities. Even though the cross-validation accuracy of such a model was poorer compared with other models developed for activities such as walking and running, as demonstrated by our results, the use of these models is crucial to achieve an acceptable prediction accuracy. It was demonstrated by testing a model previously developed by our group [@Veras2020] in a similar sample, but on walking, on the accelerometer-recorded jumping data from the current study. The accuracy indices yielded by applying these models have greatly worsened and these equations also produced some abnormal values, mainly in the pGRF model, in which it predicted negative values of force. These negative pGRF values were mainly as a result of a predictor, the second-degree polynomial of the acceleration, possessing a negative coefficient. While this coefficient captured the plateauing of pGRF as pACC magnitude augments during walking, effectively improving the model [@Veras2020] for walking, this same pattern is not present in the pGRF vs. pACC relationship during jumping, and, therefore, only impairs the accuracy of these equations when applyed to this kind of activity.

@@

Although the models presented in this study showed to be valid, some limitations ought to be considered. First, the choice of activity to elicit high-impact loading was jumping, and several combinations of jump type and height were tested. Nevertheless, the results presented here cannot be assured for other high-impact activities, or even for other types of jumps that were not tested. Second, no external sample for model validation was recruited. However, the leave-one-out cross-validation method that was used is the recommended in these situations [@Staudenmayer2012]. Third, to apply these models in accelerometer-recorded free-living physical activity data, there is a need to identify, in the acceleration signal, the periods in which the activity being performed is jumping. Therefore, methods for physical activity pattern detection should be used in combination with these models.
